|

### 2021-09-15 paper.Rmd
some progress

---
title: "Emotional framing and the effectiveness of corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]"
author: |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
# abstract-template: Everyone agrees that this issue is really important. But we do not know much about this specific question, although it matters a great deal, for these reasons. We approach the problem from this perspective. Our research design focuses on these cases and relies on these data, which we analyze using this method. Results show what we have learned about the question. They have these broader implications.

@@

@@

To investigate the direct effects of misinformation and corrective information conditional on

The study will be conducted as a survey experiment embedded within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. 
%The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample. Households that did not posses the technical devices to participate online have been equipped with such devices. Survey waves are conducted on a bi-monthly basis, the fieldwork for this study will take place in May 2021.
Because data collection is done by a third party, I do not have control over the sample size. From results of previous waves I expect a sample size of approximately 4,400 respondents.
Fieldwork will be conducted in May 2021. Preliminary data for my own analyses will be available in July 2021 and a scientific use file will be made publicly available at GESIS data archive in December 2021.
Pre-registered at OSF: https://osf.io/y63xu

@@

## Ethics {-}

@@

This paper is written in RMarkdown based on a template available here: https://osf.io/q395s

>>>>

---
title: |
  | Emotional framing and the effectiveness of 
  | corrective information^[I thank Thomas Bräuninger, Sebastian Juhl, the GIP team, and participants of the Fall 2020 CDSS workshop in political science at the University of Mannheim for excellent comments and helpful suggestions on this study. This work was supported by the German Research Foundation (DFG) via the Collaborative Research Center (SFB) 884 on ``The Political Economy of Reforms'' (Project C2) and the University of Mannheim’s Graduate School of Economics and Social Sciences.]
  | *Working Paper*
author: |
  | *Working paper*
  |
  | Pirmin Stöckle^[Corresponding address: pirmin.stoeckle@gess.uni-mannheim.de]
  | GESS and SFB 884, University of Mannheim
date: \today
linestretch: 1.5
colorlinks: true
abstract: \setstretch{1} \noindent Concerns about various forms of misinformation and its fast dissemination through online media have generated huge interest into ways to effectively correct false claims. An under-explored mechanism in this research is the role of distinct emotions. How do emotional appeals interact with corrective information? Specifically, I focus on the emotion of disgust, which has been shown to be linked to the moralization of attitudes, which in turn reduces the impact of empirical evidence on attitudes and makes compromise less likely. Substantively, I investigate the issue of genetically modified (GM) food. I implement a pre-registered survey experiment within a panel study based on a probability sample of the general population in Germany (N = 4,000). Results show what we have learned about the question. This study provides evidence on the effect of emotionally charged disinformation on perceptions of GM food, and ways to effectively correct false claims. In a broader perspective, these results inform further studies and policy interventions on other issues where disinformation loads on strong emotions, ranging from social policy over immigration to health interventions such as vaccinations. \vspace{.8cm}
# abstract-template: Everyone agrees that this issue is really important. But we do not know much about this specific question, although it matters a great deal, for these reasons. We approach the problem from this perspective. Our research design focuses on these cases and relies on these data, which we analyze using this method. Results show what we have learned about the question. They have these broader implications.

@@

<!-- This paper is written in RMarkdown based on a template available here: https://osf.io/q395s -->

@@

To investigate the direct effects of misinformation and corrective information conditional on emotional framing, I conducted a pre-registered survey experiment.^[Pre-registration can be found on OSF: https://osf.io/y63xu] The experiment was implemented within wave 53 of the German Internet Panel (GIP) [@blom2015setting]. The GIP is a longitudinal panel survey based on a probability sample of the general population in Germany. The survey is conducted online, but recruitment was conducted offline according to a strict random probability sample. Households that did not posses the technical devices to participate online have been equipped with such devices. Fieldwork for this study took place in May 2021 and the sample size reached 4,260 respondents.

The between-subjects survey experiment followed a 3 $\times$ 3 full factorial design varying both misinformation (none, low emotional framing, high emotional framing) and corrective information (none, neutral correction, counterframe correction). Respondents were randomly assigned to one of the 9 resulting groups, thus yielding 8 treatment conditions and one control condition were no information treatment was shown.

@@

@@
